---
title: "Download ALA data"
author: "Fonti Kar"
date: "2023-05-25"
output: html_document
---

Code to download occurrence records from ALA using `{galah}` using a user-supplied species list. In this case, a species list from the naming authority - Australian Fauna Directory (AFD). 

**Note: to avoid parallel process this step to avoid crashes to API system**

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      eval = FALSE)
```

### Install and load packages we need

```{r}
# For CoordinateCleaner to currently work!  Uncomment and install once
# devtools::install_github("ropensci/rnaturalearth")
# devtools::install_github("ropensci/rnaturalearthdata")
# devtools::install_github("ropensci/rnaturalearthhires")

# install.packages("pacmac")
pacman::p_load(galah, dplyr, CoordinateCleaner, sf, ozmaps, job, arrow, purrr, data.table, ggplot2)
```

### Configure ALA account

```{r}
galah_config(email = Sys.getenv("ALA_EMAIL"), # Put your email here or use usethis::usethis::edit_r_environ() to save email as enviroment variable
             atlas = "Australia",
             download_reason_id = 10 # 10 = Testing
             )
```

### Reading in the AFD species list

The list is requested from AFD and is in `.csv` format. Unfortunately, this file type makes it too large (66.2MB) it for GitHub. Here I have saved it as a `.parquet` thereby reducing its size to 17.3MB so that this workflow is more reproducible. One benefit of using `.parquet` is that the data needed is not read into memory until you call `collect()`.

```{r}
# afd_taxonomy <- fread("output/afd_splist_full.csv")
# afd_taxonomy |> write_parquet("data/AFD_splist")

afd_phylum <- open_dataset("data/AFD_splist") |> 
  select(PHYLUM) |> 
  distinct() |> 
  rename(phylum = PHYLUM) |>  #Rename phylum to lower case for search_taxa() 
  collect()

nrow(afd_phylum)
```

### Obtain count of number of occurence records for each Invertebrate Phylum  and write as .csv 

Fields for [`basisOfRecord`](https://github.com/AtlasOfLivingAustralia/ala-dataquality/wiki/basisOfRecord)

**Note:** ALA has collapsed `EnvironmentalDNA`, `GenomicDNA`" into `Material_Sample` 
ALA has added [`contentTypes`](https://github.com/AtlasOfLivingAustralia/ala-dataquality/wiki/contentTypes) to further delineate these categories

```{r}
results <- search_taxa(afd_phylum) # Issues with ACANTHOCEPHALA its technically a Phylum but recorded as a class in here https://bie.ala.org.au/species/NZOR-6-53977

# Obtain counts for each phylum, equivalent to n.all 
n.all <- galah_call() |> 
  galah_identify(results) |> 
  group_by(phylum) |> 
  atlas_counts() 

# To view more rows
n.all |> print(n = 50) # Change n for view more

# Obtain counts for human observations for each phylum, equivalent to n.obs
n.obs <- galah_call() |>
  galah_identify(results) |> 
  galah_filter(
    basisOfRecord == "HUMAN_OBSERVATION" # Filters to human observed records
  )  |> 
  galah_group_by(phylum) |> 
  atlas_counts() 

n.obs |> print(n = 50)

# Obtain counts for records that are associated with a specimen by each phylum, equivalent to n.spec
specimen_only <- c("PRESERVED_SPECIMEN", "LIVING_SPECIMEN", 
                   "MACHINE_OBSERVATION", "MATERIAL_SAMPLE")

n.spec <- galah_call() |>
  galah_identify(results) |> 
  galah_filter(
    basisOfRecord == specimen_only #  Filters to specimen only
  )  |> 
  galah_group_by(phylum)|> 
  atlas_counts()  

# Format data to save as a .csv
# Rename
n.all <- n.all |> rename(n.all = count)
n.obs <- n.obs |> rename(n.obs = count)
n.spec <- n.spec |> rename(n.spec = count) 

# Join all counts
phylum_counts <- left_join(n.all, n.obs) |> 
  left_join(., n.spec) 

# Save as .csv
# write.csv(taxon.counts, "output/taxon_counts.csv")
```

### Filter query by assertions

Assertions are used to filter down records ensure the data is fit for use. 

To display all assertion fields use:

```{r}
assertions <- show_all("assertions") |> print(n = 250)

# Save assertion data as .csv
# write_csv(assertions, "output/assertions.csv")
```

Users can filter records using `galah_filter()`
This method can also applied to downloading records as well as obtaining counts as we did above for `basisOfRecord`

Note that assertions are logicals (TRUE/FALSE) and are generally framed negatively e.g `identificationIncorrect`, TRUE is where identification was incorrect and FALSE is where identification was NOT incorrect

Below, replace`identificationIncorrect` with whatever assertion name you’d like to use. 

Alternatively, to exclude by multiple assertions, use `c()` 

```{r}
# For single assertion exclusions
galah_call() |>
  galah_identify(results) |> 
  galah_filter(
    basisOfRecord == specimen_only, # Filters to specimen only
    identificationIncorrect == FALSE # Keep correctly identified records
  ) |> 
  galah_group_by(phylum) |> 
  atlas_counts() 

# For exclusions using multiple assertions
IA_assertions <- c("UNKNOWN_KINGDOM", "identificationIncorrect",
                   "COORDINATE_PRECISION_MISMATCH", "MISSING_GEODETICDATUM")

galah_call() |>
  galah_identify(results) |> 
  galah_filter(
    basisOfRecord == specimen_only, # Filters to specimen only
    assertions != IA_assertions
  ) |> 
  galah_group_by(phylum) |> 
  atlas_counts()
```

#### Download occurrence records with assertion fields

To download assertion fields with occurrence records using `galah_select()`. Note that due changes to the ALA API systems means that there are limits to query size so this approach is recommended for smaller queries (e.g for a particular year == 2022). 

Alternatively, you can select for specific assertions by specifying their name in 
`galah_select()`
```{r}
# Getting all assertions for a small query
inverts_with_assertions <- galah_call() |>
  galah_identify(results) |> 
  galah_filter(
    basisOfRecord == specimen_only, # Filters to specimen only
    identificationIncorrect == FALSE, # Keep correctly identified records
    year == 2022)  |> # Limiting to 2022 for now
  galah_select(group = c("basic", "assertions")) |>
  atlas_occurrences() 

# Getting a few selected assertions for a larger query
# IA_assertions <- c("UNKNOWN_KINGDOM", "COORDINATE_PRECISION_MISMATCH", "MISSING_GEODETICDATUM")
# 
# inverts_with_selected_assertions <- galah_call() |>
#   galah_identify("ANNELIDA") |> 
#   galah_filter(
#     basisOfRecord == "HUMAN_OBSERVATION", 
#     identificationIncorrect == FALSE, 
#     year == 2000)  |> 
#   galah_select(group = "basic", IA_assertions) |>
#   atlas_occurrences() 
```

### Extra fields for a `galah` query

Note some fields have been renamed. Unable to find equivalent fields for: 

- `taxonomic_kosher`

Some fields did not have an exact match but here are my best guess to find an approximate match i.e. original ~ best guess

-raw_taxon_name ~ raw_scientificName
-raw_datum ~ raw_geodeticDatum
-taxon_name ~ scientificName
-common_name ~ vernacularName
-rank ~ taxonRank
-state ~ locality
-min_elevation_d not sure what _d represents ~ minimumElevationInMeters
-max_elevation_d not sure what _d represents ~ maximumElevationInMeters
-min_depth_d not sure what _d represents  ~  minimumDepthInMeters
-max_depth_d not sure what _d represents ~ maximumDepthInMeters
-collector ~ recordedBy
-occurrence_date ~ eventDate
-geospatial_kosher ~ spatiallyValid

#### Fields in ALA

Since the list of fields needed for the IA project exceeds the allowable query length of the ALA Biocache API, I have split the fields into 3 parts for the queries. These are subsequently joined back together.

```{r}
# Show all fields
ALA_fields <- show_all(fields)

ALA_fields |> print(n = 100)

# Search fields
search_all(fields, "coordinate") # Searching for fields containing coordinate
```

```{r}
# Fields that IA want
IA_fields<- c("recordID","dataResourceUid","dataResourceName",
              "institutionID","institutionName",
              "collectionID", "collectionUid","collectionName",
              "contentTypes", 
              "license", 
              "taxonConceptID",
              "raw_scientificName" ,"raw_vernacularName", 
              "scientificName", 
              "vernacularName", 
              "taxonRank", 
              "kingdom","phylum","class","order", 
              "family","genus","species","subspecies",
              "institutionCode","collectionCode",
              "locality", 
              "raw_geodeticDatum", 
              "raw_decimalLatitude","raw_decimalLongitude", 
              "decimalLatitude","decimalLongitude",
              "coordinatePrecision","coordinateUncertaintyInMeters",
              "country","stateProvince", 
              "cl959","cl21","cl1048",
              "minimumElevationInMeters", "maximumElevationInMeters", 
              "minimumDepthInMeters", "maximumDepthInMeters",
              "individualCount",
              "recordedBy",
              "eventDate",
              "year","month",
              "verbatimEventDate",
              "basisOfRecord","raw_basisOfRecord",
              "occurrenceStatus",
              "raw_sex", "sex",
              "preparations",
              "outlierLayer",
              "spatiallyValid", "catalogNumber") 

length(IA_fields)

# Split into 2 parts
IA_fields_split <- split(IA_fields, ceiling(seq_along(IA_fields)/20))

# We want to eventually join by recordID so need to add this to the 2-4 parts
IA_fields_split$`2` <- c("recordID", IA_fields_split$`2`)
IA_fields_split$`3` <- c("recordID", IA_fields_split$`3`)

# Sub-downloads for each part of the fields
 invert_2022_p1 <- galah_call() |>
  galah_identify(results) |> 
  galah_filter(basisOfRecord == specimen_only,
               year == 2022, 
               identificationIncorrect == FALSE) |>  
  galah_select(all_of(IA_fields_split$`1`)) |>
  atlas_occurrences()

invert_2022_p2 <- galah_call() |>
  galah_identify(results) |> 
  galah_filter(basisOfRecord == specimen_only,
               year == 2022, 
               identificationIncorrect == FALSE) |>  
  galah_select(all_of(IA_fields_split$`2`)) |> 
  atlas_occurrences() 

invert_2022_p3 <- galah_call() |>
  galah_identify(results) |> 
  galah_filter(basisOfRecord == specimen_only,
               year == 2022, 
               identificationIncorrect == FALSE) |>  
  galah_select(all_of(IA_fields_split$`3`)) |> 
  atlas_occurrences() 
```

#### Join the different parts together

```{r}
# Left join these together
# 2022 version
colnames(invert_2022_p1); ncol(invert_2022_p1)
colnames(invert_2022_p2); ncol(invert_2022_p2)
colnames(invert_2022_p3); ncol(invert_2022_p3)

length(IA_fields)

# 2022 version
invert_2022 <- left_join(invert_2022_p1, invert_2022_p2, by = "recordID") |> 
  left_join(invert_2022_p3, by = "recordID") 

write_parquet(invert_2022, "output/invert_2022")
```

#### RStudio job of query with required fields 

```{r}
job({
invert_2022_p1 <- galah_call() |>
  galah_identify(results) |> 
  galah_filter(basisOfRecord == specimen_only,
               year == 2022, 
               identificationIncorrect == FALSE) |>  
  galah_select(all_of(IA_fields_split$`1`)) |>
  atlas_occurrences()

write_parquet(invert_2022_p1, "output/invert_2022_p1")

invert_2022_p2 <- galah_call() |>
  galah_identify(results) |> 
  galah_filter(basisOfRecord == specimen_only,
               year == 2022, 
               identificationIncorrect == FALSE) |>  
  galah_select(all_of(IA_fields_split$`2`)) |> 
  atlas_occurrences() 

write_parquet(invert_2022_p2, "output/invert_2022_p2")

invert_2022_p3 <- galah_call() |>
  galah_identify(results) |> 
  galah_filter(basisOfRecord == specimen_only,
               year == 2022, 
               identificationIncorrect == FALSE) |>  
  galah_select(all_of(IA_fields_split$`3`)) |> 
  atlas_occurrences() 

write_parquet(invert_2022_p3, "output/invert_2022_p3")
})
```

#### Download as an arrow::parquet fie using purrr::map

I've set this is a background RStudio job for the download. `Sys.sleep(600)` is for taking a 10 minute break so to not overload the API

```{r}
# Fields that IA want
y <- tibble(name = 
                   c("recordID",
                     "dataResourceUid",
                     "dataResourceName",
                     "institutionID",
                     "institutionName",
                     "collectionID", 
                     "collectionUid",
                     "collectionName",
                     "contentTypes", 
                     "license", 
                     "taxonConceptID",
                     "raw_scientificName",
                     "raw_vernacularName", 
                     "scientificName", 
                     "vernacularName", 
                     "taxonRank", 
                     "kingdom",
                     "phylum",
                     "class",
                     "order",
                     "family",
                     "genus",
                     "species",
                     "subspecies",
                     "institutionCode",
                     "collectionCode",
                     "locality", 
                     "raw_geodeticDatum", 
                     "raw_decimalLatitude",
                     "raw_decimalLongitude", 
                     "decimalLatitude",
                     "decimalLongitude",
                     "coordinatePrecision",
                     "coordinateUncertaintyInMeters",
                     "country",
                     "stateProvince", 
                     "cl959",
                     "cl21",
                     "cl1048",
                     "minimumElevationInMeters", 
                     "maximumElevationInMeters", 
                     "minimumDepthInMeters",
                     "maximumDepthInMeters",
                     "individualCount",
                     "recordedBy",
                     "eventDate",
                     "year",
                     "month",
                     "verbatimEventDate",
                     "basisOfRecord",
                     "raw_basisOfRecord",
                     "occurrenceStatus",
                     "raw_sex", 
                     "sex",
                     "preparations",
                     "outlierLayer",
                     "spatiallyValid", 
                     "catalogNumber"),
                 type = "field") 

# Set the call as galah_select
attr(y, "call") <- "galah_select"

# Split fields into 3 parts for download
y_ls <- split(y, f = ceiling(seq_along(y$name)/20))

# Add id in 2nd and 3rd list so output can all be joined later by id
y_ls$`2` <- bind_rows(tibble(name = "recordID",
                             type = "field"),
                      y_ls$`2`)

y_ls$`3` <- bind_rows(tibble(name = "recordID",
                             type = "field"),
                      y_ls$`3`)


# The rest of the galah_call as a function
get_occ <- function(y){
  
  # Phylums
  afd_phylum <- tibble(phylum = c("ANNELIDA",
                                         "ARTHROPODA",
                                         "BRACHIOPODA",
                                         "BRYOZOA",
                                         "CHAETOGNATHA",
                                         "GASTROTRICHA",
                                         "GNATHOSTOMULIDA",
                                         "MOLLUSCA",       
                                         "NEMATODA",        
                                         "NEMATOMORPHA",    
                                         "NEMERTEA",       
                                         "ONYCHOPHORA",    
                                         "PLATYHELMINTHES",
                                         "PORIFERA",        
                                         "ROTIFERA",        
                                         "TARDIGRADA")
  )
  
  # basisOfRecord filters
  specimen_only <- c("PRESERVED_SPECIMEN", "LIVING_SPECIMEN", 
                   "MACHINE_OBSERVATION", "MATERIAL_SAMPLE")
  x <- 
  galah_call() |>
  galah_identify(afd_phylum) |> 
  galah_identify("https://biodiversity.org.au/afd/taxa/3cbb537e-ab39-4d85-864e-76cd6b6d6572",  # seperate galah_identify for acantocephala as it is a homonmym
                 search = FALSE) |>
  galah_filter(basisOfRecord == specimen_only, 
               identificationIncorrect == FALSE)
  
  x$select <- y
  
   x |> 
    atlas_occurrences() |> 
    arrow::write_parquet(sink = paste0("data/galah/IA_", Sys.time()))
   
   Sys.sleep(600)
}

## Background R Studio Job for all lists (errors - can't identify afd_taxon_phylums)
job({
  map(y_ls,
    get_occ)
})
```

#### Read in parquet and join 4 parts together

Joining the different parquets together and then save as a mass download

```{r}
# This opens the meta-data of each part
p1 <- open_dataset("data/galah/IA_2023-01-09 12:24:35")
p2 <- open_dataset("data/galah/IA_2023-01-09 12:41:56")
p3 <- open_dataset("data/galah/IA_2023-01-09 12:57:59")

# Calling the object will show you what columns are in here
p1
p2

# Joining by recordID
p1 |> 
  left_join(p2, by = "recordID") |> 
  left_join(p3, by = "recordID") |> 
  write_parquet("output/IA_mass_download")
```

## Working with parquets

```{r}
IA_data <- open_dataset("output/IA_mass_download")
IA_data

# Obtain several key variables for cleaning occurrence records
# use collect() to load into memory once you have selected what you
IA_1940onward <- IA_data |> 
  select(recordID, scientificName, starts_with("decimal"), year) |>
  filter(year >1940) |> 
  collect()

write_parquet(IA_1940onward, "output/IA_1940onwards") 
```

## To get the assertions for the bulk download

We have to break it down for each phylum as the query gets too big! The query for arthropods are separate too there is over 3 mil records to retrieve. Can be used as a background job!

Getting all the assertions for all records may not be so feasible. Perhaps limiting to a set amount of years, first deciding which assertions are important then running the query. The idea is similar to above with required fields and joining by `recordID`

Update: Not the best use of time to try get all the assertions for every record. Instead IA will preselect the required assertions and then formulate their query

```{r}
# Try mapping for each taxa
# Phylums
afd_phylum <- tibble(phylum = c("ANNELIDA",
                                "ARTHROPODA",
                                "BRACHIOPODA",
                                "BRYOZOA",
                                "CHAETOGNATHA",
                                "GASTROTRICHA",
                                "GNATHOSTOMULIDA",
                                "MOLLUSCA",       
                                "NEMATODA",        
                                "NEMATOMORPHA",    
                                "NEMERTEA",       
                                "ONYCHOPHORA",    
                                "PLATYHELMINTHES",
                                "PORIFERA",        
                                "ROTIFERA",        
                                "TARDIGRADA")
)
                     

get_occ <- function(taxa){

  # basisOfRecord filters
  specimen_only <- c("PRESERVED_SPECIMEN", "LIVING_SPECIMEN", 
                     "MACHINE_OBSERVATION", "MATERIAL_SAMPLE")
  
  x <- 
    galah_call() |>
    galah_identify("https://biodiversity.org.au/afd/taxa/3cbb537e-ab39-4d85-864e-76cd6b6d6572") |> 
    galah_filter(basisOfRecord == specimen_only, 
                 identificationIncorrect == FALSE) |> 
    galah_select(group = "assertions")

  
  x |> 
    atlas_occurrences() |> 
    arrow::write_parquet(sink = paste0("data/galah/IA_ACANTOCEPHALA"))
  
 Sys.sleep(600)
}

 map(afd_taxon_phylums,
     get_occ)

```
#### Acantocephala 

Needs to queried seperately at the moment 
```{r}
 specimen_only <- c("PRESERVED_SPECIMEN", "LIVING_SPECIMEN", 
                   "MACHINE_OBSERVATION", "MATERIAL_SAMPLE")
  
  x <- 
  galah_call() |>
  galah_identify("https://biodiversity.org.au/afd/taxa/3cbb537e-ab39-4d85-864e-76cd6b6d6572") |> 
  galah_filter(basisOfRecord == specimen_only, 
               identificationIncorrect == FALSE) |> 
    galah_select(group = "assertions")
  
   x |> 
    atlas_occurrences() |> 
    arrow::write_parquet(sink = paste0("data/galah/IA_ACANTOCEPHALA"))
```

#### Arthropoda 

Needs to be split by year or another categorical variable. Currently implementing but want to touch base first.

```{r}
specimen_only <- c("PRESERVED_SPECIMEN", "LIVING_SPECIMEN", 
                   "MACHINE_OBSERVATION", "MATERIAL_SAMPLE")
  
galah_call() |>
  galah_identify("ARTHROPODA") |>
  galah_filter(basisOfRecord == specimen_only,
               identificationIncorrect == FALSE) |>
  galah_select(group = "assertions") |>
  galah_group_by(year) |> 
  atlas_counts()
  
```

## Clean ALA records using {Coordinate Cleaner}

Check out https://docs.ropensci.org/CoordinateCleaner/

An really powerful too to remove occurrence records that are in the seas! Not a perfect solution but a great first pass

There was an issue with removing sea coordinates: https://github.com/ropensci/CoordinateCleaner/issues/77

### Precleaning

Removing missing values and d-eduplication

```{r}
#Removing missing data
IA_1940onwards |> 
  filter(is.na(decimalLongitude), is.na(decimalLatitude)) |> 
  nrow() # Removing  82176 rows

#De-duplication
IA_1940onwards |> 
  filter(duplicated(decimalLongitude), duplicated(decimalLatitude)) |> 
  nrow() # Removing   3128992 rows

IA_1940onwards_preclean <- IA_1940onwards |> 
  filter(!is.na(decimalLongitude), !is.na(decimalLatitude)) |> 
  filter(!duplicated(decimalLongitude), !duplicated(decimalLatitude)) 
```

### Removing records in the sea

```{r}
cc_flags <- IA_1940onwards_preclean |> 
  cc_sea(lon = "decimalLongitude", lat = "decimalLatitude", value = "flagged") 

summary(cc_flags)
plot(cc_flags, lon = "decimalLongitude", lat = "decimalLatitude")

# Subset sea records
IA_1940onwards_preclean |> filter(cc_flags) |> nrow()

# Exclude sea records
IA_1940onwards_searemoved <- IA_1940onwards_preclean |> filter(!cc_flags)
```
