---
title: "AFD checklist"
author: "Payal Bal, Fonti Kar"
date: "2022-08-30"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r }
#install.packages("pacman")
pacman::p_load(dplyr, janitor, galah, stringr, tidyverse, glue)

## Functions 
source("scripts/remove_improper_names_v2.R") # Version Fonti created
```

### Australian Faunal Directory taxonomy

```{r}
afd_species <- read_csv("output/afd_splist_full.csv")
all_species <- afd_species$VALID_NAME 
```

### Remove improper names 

```{r}
species_record <- remove_improper_names_v2(all_species,
                                           allow.higher.taxa = FALSE,
                                           allow.subspecies = TRUE)

# Double checking if there are any NAs in updated list
is.na(species_record$updated_list) |> table() 

# Filter records based on updated list
afd_species |> 
  dplyr::filter(VALID_NAME %in% species_record$updated_list) -> afd_species
```

### Remove invasive species 

Notes: *ALA still working to update this section on removing invasive/introduce species
Code below is written by Payal. In the next version of the `galah`, I believe we can access the GRIIS list directly through `galah`

```{r ##}
griis_species <- read.csv("data/GRIIS_Global_Register_of_Introduced_and_Invasive_Species_Australia.csv")

message(cat("number of AFD species listed in GRIIS: "),
        length(which(afd_species$VALID_NAME %in% griis_species$Supplied.Name)))

message("AFD species in GRIIS - Global Register of Introduced and Invasive Species - Australia: ")
afd_species$VALID_NAME[which(afd_species$VALID_NAME %in% griis_species$Supplied.Name)]

message("Removing AFD species in GRIIS ...")
afd_species <- afd_species[which(! afd_species$VALID_NAME %in% griis_species$Supplied.Name),]
```

### Remove Marine species

###  First option

Using CAAB data: https://www.cmar.csiro.au/data/caab/

```{r}
caab_species <- read_csv("data/caab_dump_latest.csv")

caab_species |> mutate(scientific_name = paste0(GENUS," ", SPECIES)) |> pull(scientific_name)



```


Using Interim Register of Marine and Nonmarine **Genera** https://marlin.csiro.au/geonetwork/srv/eng/catalog.search#/metadata/249c9716-3224-497e-98fc-ee0f8fd04373

```{r}
irmng_taxa <- read_csv("data/irmng_taxon.csv")
names(irmng_taxa)

irmng_taxa |> 
  filter(!is.na(genus)) |> 
  pull(genus) -> irmng_genus

# Overlap with AFD


stringr::str_detect(stringr::regex(paste(irmng_genus, collapse = "|")), all_species) |> janitor::tabyl()

```


## Identifying duplicates 

This section identify duplicates at the level of: 
- row
- `VALID_NAME`
- `COMPLETE_NAME`

The number of species > the number of _unique_ species implies that there are duplicates.
JM: use `COMPLETE_NAME` for finding duplicates


```{r}
## List duplicates comparing all columns
afd_species[duplicated(afd_species),] 

## Look for duplicates in specific columns
sum(is.na(afd_species$VALID_NAME)) # Count number of NA in VALID_NAME
length(afd_species$VALID_NAME) # Total species
length(unique(afd_species$VALID_NAME)) # Total unique VALID_NAME
length(unique(afd_species$COMPLETE_NAME)) # Total unique COMPLETE_NAME
```

## Duplicates in COMPLETE_NAME (excluding first appearance)

```{r}
message(cat("Number of duplicated COMPLETE_NAME (excluding first appearance): "), 
        length(afd_species$COMPLETE_NAME[duplicated(afd_species$COMPLETE_NAME)]))
message("duplicated COMPLETE_NAME: ")
afd_species$COMPLETE_NAME[duplicated(afd_species$COMPLETE_NAME)]
```

## Duplicates in COMPLETE_NAME (*including* first appearance)

```{r}

afd_species$COMPLETE_NAME[duplicated(afd_species$COMPLETE_NAME) | duplicated(afd_species$COMPLETE_NAME, fromLast=TRUE)] -> dup_temp

message(cat("#duplicates in COMPLETE_NAME (including first appearance) : ", length(dup_temp)))
message("duplicated COMPLETE_NAME: ")
dup_temp$COMPLETE_NAME

# readr::write_csv(temp, "output/afd_completename_repeats.csv")

```

## Resolve duplicates from ALA list 

Removal or flagging of duplicates by `TAXON_GUID` 

ALA hasn't contributed here, these were completed *in consultation with JM*

```{r}
## using TAXON_GUID from afd_completename_repeats_JRM.csv
# afd_species <- unique(afd_species$COMPLETE_NAME)
removed_dups <- c("b05771ae-bda7-497a-87c4-b55a0ebc4ca1",
                  "03acc9d4-a209-4bf0-9972-bc7d35d56aea",
                  "83d18631-e160-42ad-8332-89e4b8ba82b6",
                  "c05506f8-0188-4850-8136-7b45ea35638e",
                  "0bb19498-874f-4c6c-a637-124ec9878130")

afd_species <- afd_species[which(afd_species$TAXON_GUID %!in% removed_dups),]
#readr::write_csv(afd_species, "./outputs/afd_species_clean.csv")
```

Looks like some duplicates but not all of them are removed 5/17

These are the ones excluded:

```{r}
afd_species |> filter(TAXON_GUID %in% removed_dups) |> pull(COMPLETE_NAME)

temp$COMPLETE_NAME
```

## Various checks and tallies

Checks or tallies of species according to the number of words and particular formatting patterns in `VALID_NAME`

Note: '\' is an escape operator for various special characters

```{r}
## Counts for number of words
str_count(afd_species$VALID_NAME, pattern = "\\S+") |> janitor::tabyl()

## Counts for special characters in species names
stringr::str_count(afd_species$VALID_NAME, pattern = regex("\\\"")) |> sum() # Counting occurrences for \"
stringr::str_count(afd_species$VALID_NAME, pattern = regex("'")) |> sum() # Counting occurrences for '
stringr::str_count(afd_species$VALID_NAME, pattern = regex("\\(")) |> sum() 
stringr::str_count(afd_species$VALID_NAME, pattern = regex("\\[")) |> sum() 
```




