---
title: "Investigating WORMS"
author: "Fonti Kar"
date: "2022-08-30"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
Resources: 
https://www.marinespecies.org/aphia.php?p=webservice&type=r
https://mmisw.org/ont/ioos/marine_biogeography
```{r}
# install.packages("worrms") https://docs.ropensci.org/worrms/index.html

pacman::p_load(tidyverse, worrms, here, job)
```

Read in AFD list

```{r}
afd_taxonomy <- read_csv("output/afd_species_clean.csv")
afd_species <- unique(afd_taxonomy$VALID_NAME) 
```

## A small test

```{r}
set.seed(8)

x <- sample(afd_species, 50) 

test_alltypes_ls <- wm_records_names(x)
test_alltypes_df <- bind_rows(test_alltypes_ls)

test_df |> select(scientificname, starts_with("is")) 

x[which(x %in% marine_sp)] # Seems to work well! 

```

## Lets expand it

```{r}
# If I do full list, it runs out of memory, reducing list to 500, 200, 100 seems to work
# afd_species_ls <- wm_records_names(afd_species) |> bind_rows() 

afd_species_ls <- wm_records_names(afd_species[1:100]) |> bind_rows()
```

## Workflow to exclude marine species from AFD checklist

```{r}
afd_species_ls |> pull(scientificname) -> marine_afd

## Excluding the marine species in VALID_NAME
afd_taxonomy |> filter(! VALID_NAME %in% marine_afd) -> afd_taxonomy_nomarine
```

## Code for for loop to query their API - slow, 1hr 

```{r}
# The AFD species we want to check with WoRMS
afd_species <- unique(afd_taxonomy$VALID_NAME) 

# Split the species list to chunks of 120 species
afd_species_chunks <- split(afd_species, ceiling(seq_along(afd_species)/120))



# Create empty list to populate
worms_ls <- list()

for (i in 1:length(afd_species_chunks[1:5])) {
  
message(cat("Working on", names(afd_species_chunks)[i], "/", length(afd_species_chunks)))
  
try(worrms::wm_records_names(afd_species_chunks[[i]]) -> tmp)
    
    tmp |> bind_rows() -> worms_ls[[i]]
    
}

# Collapse the list down
worms_afd_query_results <- worms_ls %>% bind_rows()

# Save as a .csv
write_csv(worms_afd_query_results, "output/worms_afd_query_results.csv")
```

## Make the loop into a background RStudio job

This took 1hr and 6 mins

```{r}
job({
  # Create empty list to populate
worms_ls <- list()

for (i in 1:length(afd_species_chunks)) {
  
message(cat("Working on", names(afd_species_chunks)[i], "/", length(afd_species_chunks)))
  
try(worrms::wm_records_names(afd_species_chunks[[i]]) -> tmp)
    
    tmp |> bind_rows() -> worms_ls[[i]]
    
}

# Collapse the list down
worms_afd_query_results <- worms_ls %>% bind_rows()

# Save as a .csv
write_csv(worms_afd_query_results, "output/worms_afd_query_results.csv")
})
```

### Exclude marine species found by WoRMs by VALID_NAME with API query data

Options to exclude by scientificname or valid name

```{r}
worms_afd_query_results <- read_csv("output/worms_afd_query_results.csv")

# Clean the list
worms_afd_query_results %>% filter(duplicated(AphiaID)) %>% nrow()
worms_afd_query_results %>% filter(!duplicated(AphiaID)) -> worms_afd_query_results

worms_afd_query_results$scientificname |> length()
worms_afd_query_results$valid_name |> length()

(worms_afd_query_results$scientificname == worms_afd_query_results$valid_name) |> janitor::tabyl()

# Exclude the matches
afd_taxonomy |> filter(! VALID_NAME %in% worms_afd_query_results$scientificname) # 91,577
afd_taxonomy |> filter(! VALID_NAME %in% worms_afd_query_results$valid_name) # 94,147:

## The excluded taxa
afd_taxonomy |> 
  filter(VALID_NAME %in% worms_afd_query_results$scientificname) %>% 
  pull(VALID_NAME)# 21,019 species excluded


afd_taxonomy |> 
  filter(VALID_NAME %in% worms_afd_query_results$valid_name) %>% 
  pull(VALID_NAME) # 18,448  species excluded
```


## Looking at bulk download

You can request a download of WoRMS here: https://www.marinespecies.org/usersrequest.php. 
Looks like they update once a month!

No valid name just scientific name

```{r}
species <- fread(here("ignore", "WoRMS_download_2022-10-01", "speciesprofile.txt"), quote = "")
taxa <- fread(here("ignore", "WoRMS_download_2022-10-01", "taxon.txt"), quote = "")

# create a GRIIS list
worms_list <- taxa |>  
  full_join(species, by = "taxonID") 

worms_list
names(worms_list)

## Number of strictly marine species
worms_list %>%
  filter(!specificEpithet == "",
         taxonomicStatus == "accepted",
         isMarine == 1,
         isTerrestrial == 0, isFreshwater == 0, isExtinct == 0) -> accepted_marine_worms  # 57,414 specie


afd_species |> filter(VALID_NAME %in% accepted_marine_worms$scientificName) # 6,108 matched
```

