---
title: "Creating alpha-hulls in `galah`"
author: "Margot Schneider, Fonti Kar"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, 
                      eval = TRUE,
                      include = TRUE)

remotes::install_github("babichmorrowc/hull2spatial")

pacman::p_load(galah, tidyverse, alphahull, sp, sf, ozmaps, dplyr, hull2spatial, patchwork)

galah_config(email = "fonti.kar@gmail.com", 
             atlas = "Australia",
             download_reason_id = 10 # testing
             )
```

### Creating $\alpha$-hulls in `galah`

- What is convex polygon/alpha shape/alpha hull?
- What are they for in spatial analyses?
  - Estimating EOO link to Marsh et al 2021 and point to their cool plot
  - Compare areas over time link to Takach et al 2020 and point to their cool plot
- What is the distinction between convex polygons and alpha hulls
- Here's how you make with data via `galah`
  - Plot: use all records
  - Plot: Use specimen only

#### A demonstration with `iris` dataset

MARGOT: You can use these simple plots to explain what these different hulls are if you think that would help the blog! I think it is a good idea :) 

```{r}
data(iris)
iris_sepals <- iris[,1:2]

# remove duplicate datapoints
iris_sepals <- iris_sepals[!duplicated(paste(iris_sepals$Sepal.Length, iris_sepals$Sepal.Width)), ]

# a convex hull
convexhull <- chull(iris_sepals)

# create the outline of the hull
hull_pts <- c(convexhull, convexhull[1])

# plot the points and convex hull
ggplot() + 
  geom_point(data = iris_sepals, aes(Sepal.Length, Sepal.Width), colour = "darkseagreen") + 
  geom_polygon(data = iris_sepals[hull_pts, ], aes(Sepal.Length, Sepal.Width), fill = NA, colour = "black") + 
  theme_bw()
```

Here is a function to extract the polygon from an `ashape` object from the `alphahull` R package

```{r}
extract_polygons <- function(alpha_shape_obj)
{
  if(class(alpha_shape_obj) != "ashape") stop("extract_polygons requires an ashape")
  edge.df <- as.data.frame(alpha_shape_obj$edges)
  groups <- ns <- xs <- ys <- numeric(nrow(edge.df))
  m <- cbind(edge.df[[1]], edge.df[[2]])
  group <- 1
  repeat {
    i <- which(groups == 0)[1]
    if (length(i) == 0 | is.na(i)) break()
    j <- n <- 1
    repeat {
      groups[i] <- group
      ns[i] <- n
      if(j == 1) xs[i] <- edge.df$x1[i] else xs[i] <- edge.df$x2[i]
      if(j == 1) ys[i] <- edge.df$y1[i] else ys[i] <- edge.df$y2[i]
      next_ind <- which((m[, j] == m[i, j] | m[, (j %% 2 + 1)] == m[i, j]) & groups == 0)
      if (length(next_ind) == 0) break()
      j <- which(m[next_ind,] == m[i, j]) %% 2 + 1
      i <- next_ind
      n <- n + 1
    }
    group <- group + 1
  }
  data.frame(x = xs, y = ys, group = as.factor(groups))[order(groups, ns), ]
}
```


```{r}
# Create alpha shapes
alphashape_0.5 <- ashape(iris_sepals, alpha = 0.5)
alphashape_1 <- ashape(iris_sepals, alpha = 1)
alphashape_2 <- ashape(iris_sepals, alpha = 2)

# Extract polgyons 
a_0.5<- extract_polygons(alphashape_0.5)
a_1<- extract_polygons(alphashape_1)
a_2<- extract_polygons(alphashape_2)

a05 <- ggplot() + 
  geom_point(data = iris_sepals, aes(Sepal.Length, Sepal.Width), colour = "darkseagreen") + 
  geom_polygon(data = a_0.5, aes(x, y), fill = NA, colour = "black") + 
  labs(title = "a = 0.5") +
  theme_bw()

a1 <- ggplot() + 
  geom_point(data = iris_sepals, aes(Sepal.Length, Sepal.Width), colour = "darkseagreen") + 
  geom_polygon(data = a_1, aes(x, y), fill = NA, colour = "black") + 
  labs(title = "a = 1") +
  theme_bw()

a2 <- ggplot() + 
  geom_point(data = iris_sepals, aes(Sepal.Length, Sepal.Width), colour = "darkseagreen") + 
  geom_polygon(data = a_2, aes(x, y), fill = NA, colour = "black") + 
  labs(title = "a = 2") +
  theme_bw()

a05 + a1 + a2
```

### Okay lets try do this `galah`

Getting all records for a species of Damsel fly

```{r}
galah_call() |> 
  galah_identify("Austroargiolestes calcaris") |> 
  galah_select(group = "basic") |> 
   atlas_occurrences() -> dfly
```

### Data cleaning steps

```{r}
# Quick tally of number of duplicates in decimalLongitude and decimalLatitude
duplicated(dfly$decimalLongitude)  |> table()
duplicated(dfly$decimalLatitude)  |> table()

# Deduplication
dfly |> 
  filter(! duplicated(decimalLongitude) & ! duplicated(decimalLatitude)) -> dfly_ddup

# Exclude NA in coordinates
dfly_ddup |>  
  filter(! is.na(decimalLongitude) & ! is.na(decimalLatitude) ) -> dfly_clean
```

## Compute and plot alpha shape of occurrences

```{r}
dfly_clean |> 
  select(decimalLongitude, decimalLatitude) |> 
  ashape(alpha = 2) -> dfly_ashape

# Extract the ashape as a polygon
extract_polygons(dfly_ashape) -> dfly_ashape_df

# Getting a polygon of Australia and set it to the specific coordinate system
aus <- st_transform(ozmaps::ozmap_country, 4326)

ggplot() + 
  geom_sf(data = aus, colour = "black", fill = "white")  + 
  geom_point(data = dfly_clean, aes(decimalLongitude, decimalLatitude), colour = "black", size = 0.8) + 
   geom_polygon(data = dfly_ashape_df, aes(x, y), fill = "orange", colour = "black", alpha = 0.5) + 
  coord_sf(xlim=c(142, 152),ylim=c(-32,-44)) +
  labs(title = "Alpha shape", x = "Longtitude (DD)", y = "Latitude (DD)") + # DD means decimal degrees here
  theme_bw() -> dfly_ashape_p

dfly_ashape_p
```

## Compute and plot alpha hull of occurrences

Alpha hulls are a bit trickier due to their arcs. We rely on an R package `hull2spatial` that converts `ahull`. We will also be relying on the R packages `sf` for plotting purposes

### Turn ALA occurrence data into sf 

```{r}
dfly_clean |> 
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326) -> dfly_sf
```

### Compute an alpha hull for our occurrences

```{r}
dfly_clean |> 
  select(decimalLongitude, decimalLatitude) |> 
  ahull(alpha = 2) -> dfly_ahull
```

### Transform our `ahull` object into a `sp` object

```{r}
hull2spatial::ahull2poly(dfly_ahull) -> dfly_sp_df
```

### Then transform `sp` object into a `sf` object

```{r}
dfly_sp_df |> 
  st_as_sf() -> dfly_sf_ahull

# Set the coordinate system using `aus` object
dfly_sf_ahull %>% st_set_crs(st_crs(aus)) -> dfly_sf_ahull
```

```{r}
ggplot() + 
  geom_sf(data = aus, colour = "black", fill = "white")  +
  geom_sf(data = dfly_sf, colour = "black", size = 0.5) +  
  geom_sf(data = dfly_sf_ahull, fill = "orange", alpha = 0.5) +
  coord_sf(xlim=c(142, 152),ylim=c(-32,-44)) +
  labs(title = "Alpha hull", x = "Longtitude (DD)", y = "Latitude (DD)") + 
  theme_bw() -> dfly_ahull_p

dfly_ahull_p
```

### Side by side of alpha shape and alpha hull

```{r}
dfly_ashape_p + dfly_ahull_p
```

## Compute alpa hulls using records that were based on specimens only

Note the huge drop in the number of records!

```{r}
# Create a vector containing specimen records only for filtering
specimen_only <- c("PRESERVED_SPECIMEN", "LIVING_SPECIMEN", 
                   "MACHINE_OBSERVATION", "MATERIAL_SAMPLE")

galah_call() |> 
  galah_identify("Austroargiolestes calcaris") |> 
  galah_filter(basisOfRecord == specimen_only) |> 
  galah_select(group = "basic") |> 
   atlas_occurrences() -> dfly_specionly
```

### Repeat the process again

```{r}
# Deduplication
dfly_specionly |> 
  filter(! duplicated(decimalLongitude) & ! duplicated(decimalLatitude)) -> dflysp_ddup

# Exclude NA in coordinates
dflysp_ddup |>  
  filter(! is.na(decimalLongitude) & ! is.na(decimalLatitude) ) -> dflysp_clean

# Turn ALA occurrence data into sf 
dflysp_clean |> 
  st_as_sf(coords = c("decimalLongitude", "decimalLatitude"), crs = 4326) -> dflysp_sf

# Compute an alpha hull for our specimen only occurrences
dflysp_clean |> 
  select(decimalLongitude, decimalLatitude) |> 
  ahull(alpha = 2) -> dflysp_ahull

# Transform our `ahull` object into a `sp` object and then into a `sf` object
hull2spatial::ahull2poly(dflysp_ahull) |> 
  st_as_sf() -> dflysp_sf_ahull

# Set the coordinate system using `aus` object
dflysp_sf_ahull %>% st_set_crs(st_crs(aus)) -> dflysp_sf_ahull

ggplot() + 
  geom_sf(data = aus, colour = "black", fill = "white")  +
  geom_sf(data = dflysp_sf, colour = "black", size = 0.5) +  
  geom_sf(data = dflysp_sf_ahull, fill = "orange", alpha = 0.5) +
  coord_sf(xlim=c(142, 152),ylim=c(-32,-44)) +
  labs(title = "Specimen only \nalpha hull", x = "Longtitude (DD)", y = "Latitude (DD)") + 
  theme_bw() -> dflysp_ahull_p

dflysp_ahull_p
```

### All of them together

```{r}
combined <- dfly_ashape_p + dfly_ahull_p + dflysp_ahull_p

combined + plot_annotation(
  title = expression(italic("Austroargiolestes calcaris"))
)
```


