---
title: "Processing ALA data"
date: "2022-08-04"
output: html_document
editor_options: 
  chunk_output_type: console
---

## Clean downloaded ALA data and summarise
Notes: Each occurrence record has a record id, we can use `ALA4R::occurrence_details` to get additional information if you need something extra later

```{r setup}
## >>>>>> PREVIOUS SCRIPT <<<<<<< ####
file.edit("")


## Set working environment ####
rm(list = ls())
gc()

# system("ps")
# system("pkill -f R")

x <- c("data.table", "rje", "stringr")
lapply(x, require, character.only = TRUE)

## Server paths
output_dir = "outputs/"
source("scripts/remove_improper_names.R")
source("scripts/get_AFDsynonyms.R")
source("scripts/get_source.R")

## Functions
'%!in%' <- function(x,y)!('%in%'(x,y))

## Load AFD taxonomic checklist
afd_taxonomy <- read.csv("../output/afd_species_clean.csv")
```

```{r}
afd_species <- sort(unique(afd_taxonomy$VALID_NAME))
message(cat("Number of species in AFD checklist: "),
        length(afd_species))
```

```{r}
## Load ALA data
ala_raw <- readRDS(file.path(output_dir, "merged_ala_2020-10-02.rds"))
```

```{r}
ala_species <- sort(unique(ala_raw$scientificName))
```

```{r}
message(cat("Number of unique species in ALA data: "),
        length(ala_species))
```

```{r}
message(cat("NAs in ALA species list: "),
        sum(is.na(ala_species)))
```

## ALA data cleaning ####
## Improper species names found in ALA data: 

```{r}
species_record <- remove_improper_names(as.character(ala_species),
                                        allow.higher.taxa = FALSE,
                                        allow.subspecies = TRUE)

message(cat("# Improper species names found in ALA data: "),
        length(species_record$improper_species))
```

## Incomplete species names found in ALA data: 

```{r}
# message("Improper species names found in ALA data: ")
# species_record$improper_species
message(cat("# Incomplete species names found in ALA data: "),
        length(species_record$incomplete_species))
```

## NAs in cleaned ALA species list:

```{r}
message(cat("NAs in cleaned ALA species list: "),
        sum(is.na(species_record$updated_list)))
```

## Duplicates in cleaned ALA species list
```{r}
message(cat("Duplicates in cleaned ALA species list: "),
        length(species_record$updated_list[duplicated(species_record$updated_list)]))


saveRDS(species_record, file = file.path(output_dir, "improper_names.rds"))
```

## Remove improper & incomplete names from ALA data ####

```{r}
ala_species <- as.character(na.omit(species_record$updated_list))

ala_dat <- ala_raw[which(ala_raw$scientificName %in% ala_species),]
```

## Checks:

```{r}
message(cat("Number of unique species in ALA data: "),
        length(ala_species))

message(cat("Number of records in raw ALA data: "),
        nrow(ala_raw))

message(cat("Number of records in cleaned ALA data: "),
        nrow(ala_dat))

message(cat("Prop of records lost in cleaning ALA data: "),
        (nrow(ala_raw) - nrow(ala_dat))/nrow(ala_raw))

message(cat("Prop of species removed: "),
        (length(unique(ala_raw$scientificName))-
           length(ala_species))/length(unique(ala_raw$scientificName)))

## Checks
length(ala_species) == length(unique(ala_dat$scientificName))
```



```{r}
rje::is.subset(ala_species, unique(ala_dat$scientificName))
```

```{r}
rje::is.subset(unique(ala_dat$scientificName), ala_species)
```

```{r}
sum(ala_species==sort(unique(ala_dat$scientificName)))
```

## Find NZ and marine species in ALA ####

```{r}

ala_species_all <- ala_species <- sort(unique(ala_dat$scientificName))
```


```{r}
# matched_data <- suppressWarnings(get_source(ala_species))
# saveRDS(matched_data, file = file.path(output_dir, "matched_data.rds"))
matched_data <- readRDS(file.path(output_dir, "matched_data.rds"))
```

## Idenitfy source of species name as recorded in ALA database ####
```{r}
unique(matched_data[,name_source])

matched_data[,.N, by = name_source]

matched_data[name_source == "APC"]$sp_names ## list APC species

matched_data[name_source == "CoL"]$sp_names ## list CoL species
```

## Remove species 

```{r}
sources <- as.data.table(read.csv(file.path(output_dir, "name_source.csv")))

message(cat("Excluding name sources: \n"),
        paste0(sources[exclude == 1]$name_source, sep = "\n"))

sources <- sources[exclude == 1]$name_source

message(cat("Number of species removed: "),
        nrow(matched_data[name_source %in% sources]))

message(cat("Number of species retained: "),
        nrow(matched_data[name_source %!in% sources]))

ala_species <- matched_data[name_source %!in% sources]$sp_names
```

## Checks

```{r}
length(ala_species)

sum(sort(ala_species_all[ala_species_all %!in% ala_species]) == sort(matched_data[name_source %in% sources]$sp_names))
```

## >> Identify (marine) habitat for species as listed in CAAB: http://www.cmar.csiro.au/caab/ ####

```{r}
unique(matched_data[,caab_habitat]) 

matched_data[,.N, by = caab_habitat]
```

## Remove species
```{r}
marine <- as.data.table(read.csv(file.path(output_dir, "caab_habitat.csv")))

message(cat("Excluding name sources: \n"),
        paste0(marine[exclude == 1]$caab_habitat, sep = "\n"))

marine <- marine[exclude == 1]$caab_habitat

message(cat("Number of species removed: "),
        nrow(matched_data[caab_habitat %in% marine]))

message(cat("Number of species retained: "),
        nrow(matched_data[caab_habitat %!in% marine]))

ala_species <- matched_data[caab_habitat %!in% marine]$sp_names
```

## Remove NZ and marine species from ALA data ####
```{r}
ala_species <- sort(ala_species)

message(cat("Number of unique species retained: "),
        length(ala_species))

ala_dat <- ala_raw[which(ala_raw$scientificName %in% ala_species),]

message(cat("Number of records in updated ALA data: "),
        nrow(ala_dat))

message(cat("Prop of records lost: "),
        (1949710 - nrow(ala_dat)) / 1949710)

message(cat("Proprotion of species removed: "),
        (62539 - length(ala_species)) / 62539)
```

## Get rid of unusable long lat vals ####
```{r}
message(cat("Number of records with unusable lat-long:"),
        nrow(ala_dat[longitude < -180 |
                       longitude > 180 |
                       latitude < -90 |
                       latitude > 90, ]))

ala_dat <- ala_dat[longitude >= -180 &
                     longitude <= 180 &
                     latitude >= -90 &
                     latitude <= 90, ]

## Checks
range(ala_dat$latitude)
range(ala_dat$longitude)
length(ala_dat[longitude < -180]$longitude)
length(ala_dat[longitude > 180]$longitude)
length(ala_dat[latitude < -90]$latitude)
length(ala_dat[latitude > 90]$latitude)
```

## Identify ALA species not found in AFD checklist ####
ALA scientificName compared against VALID_NAME and SYNONYMS in AFD checklist
Species names categorised by number of words in the name for comparisons  

```{r}
ala_species <- sort(unique(ala_dat$scientificName))

message(cat("Are all ALA species contained in the AFD checklist: "),
        rje::is.subset(ala_species, afd_species))
```

## >> ALA species: 4-word names ####
```{r}
ala_species <- sort(unique(ala_dat$scientificName))

sp_words <- sapply(strsplit(as.character(ala_species), " "), length)
unique(sp_words)
n = 4
length(ala_species[which(sp_words == n)])

message(cat("Number of species not found in AFD: "),
        sum(ala_species[which(sp_words == n)] %!in% afd_species))

ala_names <- sort(ala_species[which(sp_words == n)][ala_species[which(sp_words == n)] %!in% afd_species])

temp <- ala_dat[scientificName %in% ala_names]
write.csv(temp, file.path(output_dir, "names4_records.csv"))

rm(temp)
```

## Checking for species manually

```{r}
matches <- get_AFDsynonyms(ala_names, afd_taxonomy)

message(cat("Synonyms found for full species names for: "),
        sum(!is.na(sapply(matches, "[[", 1))), " species")

message(cat("Synonyms found for genus (only) names for: "),
        sum(!is.na(sapply(matches, "[[", 2))), " species")

message(cat("Synonyms found for species (only) names for: "),
        sum(!is.na(sapply(matches, "[[", 3))), " species")
```

## Save outputs
```{r}
y <- rbindlist(matches, fill=FALSE)
y <- cbind(ala_names, y)

write.csv(y, file.path(output_dir, paste0("names", n, ".csv")), row.names = FALSE)
```

## Remove or mark unmatched and partially matched species from dataset
Species list reviewed by JM: see names2_JRM.csv
all species were found to either be marine or from NZ

```{r}
message(cat("Number of records to be removed: "),
        length(which(ala_dat$scientificName %in% ala_names)))

## Remove records
dim(ala_dat)

ala_dat <- ala_dat[which(ala_dat$scientificName %!in% ala_names),]
dim(ala_dat)
```

## >> ALA species: 3-word names ####

```{r}
ala_species <- sort(unique(ala_dat$scientificName))

sp_words <- sapply(strsplit(as.character(ala_species), " "), length)

unique(sp_words)
n = 3
length(ala_species[which(sp_words == n)])

ala_names <- sort(ala_species[which(sp_words == n)][ala_species[which(sp_words == n)] %!in% afd_species])
```

```{r}
message(cat("Number of species not found in AFD: "),
        sum(ala_species[which(sp_words == n)] %!in% afd_species))

message(cat("Number of ALA records associated with species found: "),
        length(which(ala_dat$scientificName %in% ala_names)))

## Checking for species manually
matches <- get_AFDsynonyms(ala_names, afd_taxonomy)

message(cat("Synonyms found for full species names for: "),
        sum(!is.na(sapply(matches, "[[", 1))), " species")

```{r}
names(matches[!is.na(sapply(matches, "[[", 1))])

message(cat("Synonyms found for genus (only) names for: "),
        sum(!is.na(sapply(matches, "[[", 2))), " species")

names(matches[!is.na(sapply(matches, "[[", 2))])

message(cat("Synonyms found for species (only) names for: "),
        sum(!is.na(sapply(matches, "[[", 3))), " species")

names(matches[!is.na(sapply(matches, "[[", 3))])

## Save outputs
y <- rbindlist(matches, fill=FALSE)
y <- cbind(ala_names, y)

write.csv(y, file.path(output_dir, paste0("names", n, ".csv")), row.names = FALSE)
```

## Remove or mark unmatched and partially matched species from dataset
Species list reviewed by JM: see names3_JRM.csv
marine, NZ or exotic species idenntified were removed
```{r}
ala_names <- fread(file.path(output_dir, paste0("names", n, "_JRM.csv")))

names3_out <- ala_names[exclude == 1]$ala_names

names3_in <- ala_names[exclude == 0]$ala_names

message(cat("Total number of species checked: "),
        length(ala_names))

message(cat("Number of species to be removed: "),
        length(names3_out))
```

```
## Number of species to be removed:
```

```
## Error in lapply(list(...), as.character): object 'names3_out' not found
```

```{r}
message(cat("Number of records to be removed: "),
        length(which(ala_dat$scientificName %in% names3_out)))
```

```
## Number of records to be removed:
```

```
## Error in ala_dat$scientificName %in% names3_out: object 'ala_dat' not found
```

```{r}
message(cat("Number of species to be marked: "),
        length(names3_in))
```

```
## Number of species to be marked:
```

```
## Error in lapply(list(...), as.character): object 'names3_in' not found
```

```{r}
message(cat("Number of records to be marked for checking: "),
        length(which(ala_dat$scientificName %in% names3_in)))
```

```
## Number of records to be marked for checking:
```

```
## Error in ala_dat$scientificName %in% names3_in: object 'ala_dat' not found
```

```{r}
## Remove records
dim(ala_dat)
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
ala_dat <- ala_dat[which(ala_dat$scientificName %!in% names3_out),]
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
dim(ala_dat)
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
## Mark records
ala_dat <- cbind(ala_dat, rep(NA, nrow(ala_dat)))
```

```
## Error in cbind(ala_dat, rep(NA, nrow(ala_dat))): object 'ala_dat' not found
```

```{r}
names(ala_dat)[ncol(ala_dat)] <- paste0("names", n)
```

```
## Error in names(ala_dat)[ncol(ala_dat)] <- paste0("names", n): object 'ala_dat' not found
```

```{r}
setkey(ala_dat, scientificName)
```

```
## Error in setkey(ala_dat, scientificName): object 'ala_dat' not found
```

```{r}
ala_dat[names3_in, paste0("names", n) := 1]
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
ala_dat[scientificName %!in% names3_in, paste0("names", n) := 0]
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
## Checks
length(names3_in) == sum(sort(unique(ala_dat[names3_in]$scientificName)) == sort(names3_in))
```

```
## Error in eval(expr, envir, enclos): object 'names3_in' not found
```

```{r}
nrow(ala_dat[names3 == TRUE]) == length(which(ala_dat$scientificName %in% names3_in))
```

```
## Error in nrow(ala_dat[names3 == TRUE]): object 'ala_dat' not found
```

```{r}
## >> ALA species: 2-word names ####
ala_species <- sort(unique(ala_dat$scientificName))
```

```
## Error in unique(ala_dat$scientificName): object 'ala_dat' not found
```

```{r}
sp_words <- sapply(strsplit(as.character(ala_species), " "), length)
```

```
## Error in strsplit(as.character(ala_species), " "): object 'ala_species' not found
```

```{r}
unique(sp_words)
```

```
## Error in unique(sp_words): object 'sp_words' not found
```

```{r}
n = 2

length(ala_species[which(sp_words == n)])
```

```
## Error in eval(expr, envir, enclos): object 'ala_species' not found
```

```{r}
ala_names <- sort(ala_species[which(sp_words == n)][ala_species[which(sp_words == n)] %!in% afd_species])
```

```
## Error in sort(ala_species[which(sp_words == n)][ala_species[which(sp_words == : object 'ala_species' not found
```

```{r}
message(cat("Number of species not found in AFD: "),
        sum(ala_species[which(sp_words == n)] %!in% afd_species))
```

```
## Number of species not found in AFD:
```

```
## Error in x %in% y: object 'ala_species' not found
```

```{r}
message(cat("Number of ALA records associated with species found: "),
        length(which(ala_dat$scientificName %in% ala_names)))
```

```
## Number of ALA records associated with species found:
```

```
## Error in ala_dat$scientificName %in% ala_names: object 'ala_dat' not found
```

```{r}
## Checking for species manually
matches <- get_AFDsynonyms(ala_names, afd_taxonomy)
```

```
## Error in get_AFDsynonyms(ala_names, afd_taxonomy): could not find function "get_AFDsynonyms"
```

```{r}
message(cat("Synonyms found for full species names for: "),
        sum(!is.na(sapply(matches, "[[", 1))), " species")
```

```
## Synonyms found for full species names for:
```

```
## Error in lapply(X = X, FUN = FUN, ...): object 'matches' not found
```

```{r}
names(matches[!is.na(sapply(matches, "[[", 1))])
```

```
## Error in eval(expr, envir, enclos): object 'matches' not found
```

```{r}
message(cat("Synonyms found for genus (only) names for: "),
        sum(!is.na(sapply(matches, "[[", 2))), " species")
```

```
## Synonyms found for genus (only) names for:
```

```
## Error in lapply(X = X, FUN = FUN, ...): object 'matches' not found
```

```{r}
names(matches[!is.na(sapply(matches, "[[", 2))])
```

```
## Error in eval(expr, envir, enclos): object 'matches' not found
```

```{r}
message(cat("Synonyms found for species (only) names for: "),
        sum(!is.na(sapply(matches, "[[", 3))), " species")
```

```
## Synonyms found for species (only) names for:
```

```
## Error in lapply(X = X, FUN = FUN, ...): object 'matches' not found
```

```{r}
names(matches[!is.na(sapply(matches, "[[", 3))])
```

```
## Error in eval(expr, envir, enclos): object 'matches' not found
```

```{r}
## Save outputs
y <- rbindlist(matches, fill=FALSE)
```

```
## Error in rbindlist(matches, fill = FALSE): object 'matches' not found
```

```{r}
y <- cbind(ala_names, y)
```

```
## Error in cbind(deparse.level, ...): object 'y' not found
```

```{r}
write.csv(y, file.path(output_dir, paste0("names", n, ".csv")), row.names = FALSE)
```

```
## Error in is.data.frame(x): object 'y' not found
```

```{r}
## Record and keep species in dataset: Unmacthed & partially matched species
message(cat("Number of records to be marked for checking: "),
        length(which(ala_dat$scientificName %in% ala_names)))
```

```
## Number of records to be marked for checking:
```

```
## Error in ala_dat$scientificName %in% ala_names: object 'ala_dat' not found
```

```{r}
ala_dat <- cbind(ala_dat, rep(NA, nrow(ala_dat)))
```

```
## Error in cbind(ala_dat, rep(NA, nrow(ala_dat))): object 'ala_dat' not found
```

```{r}
names(ala_dat)[ncol(ala_dat)] <- paste0("names", n)
```

```
## Error in names(ala_dat)[ncol(ala_dat)] <- paste0("names", n): object 'ala_dat' not found
```

```{r}
setkey(ala_dat, scientificName)
```

```
## Error in setkey(ala_dat, scientificName): object 'ala_dat' not found
```

```{r}
ala_dat[ala_names, paste0("names", n) := 1]
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
ala_dat[scientificName %!in% ala_names, paste0("names", n) := 0]
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
## Check
nrow(ala_dat[names2 == TRUE]) == length(which(ala_dat$scientificName %in% ala_names))
```

```
## Error in nrow(ala_dat[names2 == TRUE]): object 'ala_dat' not found
```

```{r}
## Save cleaned ALA data ####
saveRDS(ala_dat, file = file.path(output_dir, paste0("clean1_ala_", Sys.Date(),".rds")))
```

```
## Error in saveRDS(ala_dat, file = file.path(output_dir, paste0("clean1_ala_", : object 'ala_dat' not found
```

```{r}
write.csv(ala_dat, file = file.path(output_dir, paste0("clean1_ala_", Sys.Date(),".csv")))
```

```
## Error in is.data.frame(x): object 'ala_dat' not found
```

```{r}
message(cat("Number of records in data: "),
        nrow(ala_dat))
```

```
## Number of records in data:
```

```
## Error in nrow(ala_dat): object 'ala_dat' not found
```

```{r}
## Drop qa columns without values (i.e. all FALSE values) ####
qa <- as.data.frame(read.csv(file.path(output_dir, "qa_assertions.csv")))
```

```
## Warning in file(file, "rt"): cannot open file '/Users/fontikar/Dropbox/1 - ALA/Projects/data_cleaning_workflows/
## scripts/nesp_bugs/outputs/qa_assertions.csv': No such file or directory
```

```
## Error in file(file, "rt"): cannot open the connection
```

```{r}
x <- qa$name
```

```
## Error in eval(expr, envir, enclos): object 'qa' not found
```

```{r}
drop_cols <- c()
for(i in x){
        if(eval(parse(text = paste0("sum(ala_dat$", i, ")"))) == 0){
                drop_cols <- c(drop_cols, i)
        }
}
```

```
## Error in eval(expr, envir, enclos): object 'x' not found
```

```{r}
message("Columns being dropped:")
```

```
## Columns being dropped:
```

```{r}
drop_cols
```

```
## NULL
```

```{r}
message(cat("Number of columns being dropped: "),
        length(drop_cols))
```

```
## Number of columns being dropped:
```

```
## 0
```

```{r}
ala_dat[, (drop_cols) := NULL]
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
message(cat("Number of columns in updated data: "),
        dim(ala_dat)[2])
```

```
## Number of columns in updated data:
```

```
## Error in lapply(list(...), as.character): object 'ala_dat' not found
```

```{r}
## Drop records with issues ####
## Issues as commented by JM
exclude <- qa[which(qa$exclude == 1),]$name
```

```
## Error in eval(expr, envir, enclos): object 'qa' not found
```

```{r}
out <- c()

## Find number of records with issues 
for(i in exclude){
        message(cat(paste0("Number of records with issue - ", i, ": ")),
                eval(parse(text = paste0("sum(ala_dat$", i, ")"))))
        out <- c(out, eval(parse(text = paste0("sum(ala_dat$", i, ")"))))
}
```

```
## Error in eval(expr, envir, enclos): object 'exclude' not found
```

```{r}
## Remove records
for(i in 1: length(exclude)){
        if(any(out[i] > 0)){
                ala_dat <- eval(parse(text = paste0("ala_dat[", exclude[i], " == TRUE]")))
        } else {
                message(cat("No records found with issue: "),
                        exclude[i])
        }
}
```

```
## Error in eval(expr, envir, enclos): object 'exclude' not found
```

```{r}
## Create species file names & add to data as column ####
## NOTE: This step revelas duplicates in species names in the data;
## These are corrected in the data, not dropped
ala_species <- sort(unique(ala_dat$scientificName))
```

```
## Error in unique(ala_dat$scientificName): object 'ala_dat' not found
```

```{r}
message(cat("Number of species with data: "),
        length(ala_species))
```

```
## Number of species with data:
```

```
## Error in lapply(list(...), as.character): object 'ala_species' not found
```

```{r}
message(cat("Duplicates found in ala_species: "),
        sum(duplicated(ala_species)))
```

```
## Duplicates found in ala_species:
```

```
## Error in duplicated(ala_species): object 'ala_species' not found
```

```{r}
## Text modification
spfilename <- stringr::str_replace_all(ala_species, " ", "00xx00")
```

```
## Error in stri_replace_all_regex(string, pattern, fix_replacement(replacement), : object 'ala_species' not found
```

```{r}
message(cat("Duplicates found in spfilename: "),
        sum(duplicated(spfilename)))
```

```
## Duplicates found in spfilename:
```

```
## Error in duplicated(spfilename): object 'spfilename' not found
```

```{r}
spfilename <- str_replace_all(spfilename, "[^[:alnum:]]", "")
```

```
## Error in stri_replace_all_regex(string, pattern, fix_replacement(replacement), : object 'spfilename' not found
```

```{r}
message(cat("Duplicates found in spfilename: "),
        sum(duplicated(spfilename)))
```

```
## Duplicates found in spfilename:
```

```
## Error in duplicated(spfilename): object 'spfilename' not found
```

```{r}
spfilename <- tolower(gsub("00xx00", "_", spfilename))
```

```
## Error in is.factor(x): object 'spfilename' not found
```

```{r}
message(cat("Duplicates found in spfilename: "),
        sum(duplicated(spfilename)))
```

```
## Duplicates found in spfilename:
```

```
## Error in duplicated(spfilename): object 'spfilename' not found
```

```{r}
spfilename <- unique(spfilename)
```

```
## Error in unique(spfilename): object 'spfilename' not found
```

```{r}
length(unique(spfilename))
```

```
## Error in unique(spfilename): object 'spfilename' not found
```

```{r}
## Identify duplicates in species name in data
spfilename[duplicated(spfilename)]
```

```
## Error in eval(expr, envir, enclos): object 'spfilename' not found
```

```{r}
grep("incertae", spfilename)
```

```
## Error in is.factor(x): object 'spfilename' not found
```

```{r}
ala_species[grep("incertae", spfilename)]
```

```
## Error in eval(expr, envir, enclos): object 'ala_species' not found
```

```{r}
temp <- ala_species[grep("incertae", spfilename)]
```

```
## Error in eval(expr, envir, enclos): object 'ala_species' not found
```

```{r}
dim(ala_dat[scientificName == temp[1]])
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
dim(ala_dat[scientificName == temp[2]])
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
## Modify data to correct duplicates in species name; no records dropped
ala_dat[scientificName == temp[2], scientificName := temp[1]]
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
dim(ala_dat[scientificName == temp[1]]) ## check
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
dim(ala_dat[scientificName == temp[2]]) ## check
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
## Add species file names column to data
y <- ala_dat$scientificName
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
y <- str_replace_all(y, " ", "00xx00")
```

```
## Error in stri_replace_all_regex(string, pattern, fix_replacement(replacement), : object 'y' not found
```

```{r}
y <- str_replace_all(y, "[^[:alnum:]]", "")
```

```
## Error in stri_replace_all_regex(string, pattern, fix_replacement(replacement), : object 'y' not found
```

```{r}
y <- tolower(gsub("00xx00", "_", y))
```

```
## Error in is.factor(x): object 'y' not found
```

```{r}
x <- sort(unique(y))
```

```
## Error in unique(y): object 'y' not found
```

```{r}
sum(x == sort(spfilename)) ## Check
```

```
## Error in eval(expr, envir, enclos): object 'x' not found
```

```{r}
ala_dat[, spfile := y]
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
dim(ala_dat)
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
## Display fields used to identify duplicates & check for NAs within ####
grep("ID|id", names(ala_dat), value = TRUE)
```

```
## Error in is.factor(x): object 'ala_dat' not found
```

```{r}
grep("catalogue", names(ala_dat), value = TRUE)
```

```
## Error in is.factor(x): object 'ala_dat' not found
```

```{r}
sum(is.na(ala_dat$collectionID)) ## https://dwc.tdwg.org/list/#dwc_collectionID
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
sum(is.na(ala_dat$institutionID))
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
sum(is.na(ala_dat$catalogueNumber))
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
## Check 'duplicateStatus' fields - NOT USEFUL
## This field is not used to subset data
grep("dup", names(ala_dat), value = TRUE)
```

```
## Error in is.factor(x): object 'ala_dat' not found
```

```{r}
unique(ala_dat$duplicateStatus) ## https://github.com/AtlasOfLivingAustralia/ala-dataquality/wiki/duplicate_status
```

```
## Error in unique(ala_dat$duplicateStatus): object 'ala_dat' not found
```

```{r}
ala_dat[,.N,by = duplicateStatus]
```

```
## Error in eval(expr, envir, enclos): object 'ala_dat' not found
```

```{r}
## Summary of cleaned data ####
message(cat("Number of records in cleaned data: "),
        nrow(ala_dat))
```

```
## Number of records in cleaned data:
```

```
## Error in nrow(ala_dat): object 'ala_dat' not found
```

```{r}
ala_species <- sort(unique(ala_dat$scientificName))
```

```
## Error in unique(ala_dat$scientificName): object 'ala_dat' not found
```

```{r}
message(cat("Number of unique species in cleaned data: "),
        length(ala_species))
```

```
## Number of unique species in cleaned data:
```

```
## Error in lapply(list(...), as.character): object 'ala_species' not found
```

```{r}
## Save updated data (all) ####
saveRDS(ala_dat, file = file.path(output_dir, paste0("clean2_ala_", Sys.Date(),".rds")))
```

```
## Error in saveRDS(ala_dat, file = file.path(output_dir, paste0("clean2_ala_", : object 'ala_dat' not found
```

```{r}
write.csv(ala_dat, file = file.path(output_dir, paste0("clean2_ala_", Sys.Date(),".csv")))
```

```
## Error in is.data.frame(x): object 'ala_dat' not found
```

```{r}
## >>>>>> NEXT SCRIPT <<<<<<< ####
file.edit("/tempdata/workdir/nesp_bugs/scripts/nonala_processing.R")
```

