---
title: "Invert Multispecies Alpha Hulls"
author: "Fonti Kar"
date: "2023-01-19"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(remotes, galah, dplyr, tidyverse,alphahull, sp, sf, ozmaps, hull2spatial, patchwork, here, stringr, skimr)

aus <- st_transform(ozmaps::ozmap_country, 4326)
source("ignore/ahull.R")
```


### Configuring ALA account
```{r}
galah_config(email = "fonti.kar@gmail.com", 
             atlas = "Australia",
             download_reason_id = 10 # testing reason
             )
```

### Read in data

```{r}
inverts <- read_csv(here("data", "inverts", "invertebrate.data.fixed.csv"))

str(inverts)
skim(inverts)

```

### Preclean for alpha hulls

```{r}
clean_inv <- inverts %>% 
  filter(! duplicated(longitude) & ! duplicated(latitude)) %>%  # Exclude duplicated coords
  drop_na(longitude, latitude) # Exclude records with missing coords

inverts %>% 
  filter(duplicated(longitude) & duplicated(latitude)) %>% nrow() # This is a massive amount of data
```

### Summaries of all taxonomic levels

Plot by class might be better! Each family is only represented by 1 taxa

```{r}
clean_inv %>% 
  group_by(class) %>% 
  summarise(n_sp = length(unique(scientific_name))) %>% 
  arrange(-n_sp)  %>% 
  print(n = 50) 

clean_inv %>% 
  group_by(family) %>% 
  summarise(n_sp = length(unique(scientific_name))) %>%
  arrange(-n_sp) 

## Other class
other_class <- clean_inv %>% 
  group_by(class) %>% 
  summarise(n_sp = length(unique(scientific_name))) %>%
  filter(n_sp <= 10 | is.na(class)) %>% 
  pull(class)

## Case study
clean_inv %>% 
  filter(class == "Arachnida", 
         family == "Salticidae") %>% 
  group_by(scientific_name) %>% 
  summarise(n()) %>% 
  arrange(-`n()`) %>% slice(1:10) %>% 
  pull(scientific_name) -> top10jspiders #Top 5 species
  
``` 

### Create new class variable

```{r}
clean_inv <- clean_inv %>% 
  mutate(plot_class = ifelse(class %in% other_class, "Other", class)) 

clean_inv %>% 
  group_by(plot_class) %>% 
  summarise(n_sp = length(unique(scientific_name))) %>% 
  arrange(-n_sp)  %>% 
  print(n = 50) 
```

### Exclude species with fewer than 4 records

```{r}
too_few_taxa <- clean_inv %>% 
  group_by(scientific_name) %>% 
  summarise(n = n()) %>% 
  filter(n < 4) %>% 
  pull(scientific_name)

# Exclude too_few_taxa
plotdat_inv <- clean_inv %>% 
  filter(!scientific_name %in% too_few_taxa)
```

### Compile data into tibble 

```{r}
nest_inv <- plotdat_inv %>%
  dplyr::select(family, plot_class, scientific_name, longitude, latitude) %>%
  group_by(family, plot_class, scientific_name) %>%
  arrange(family, plot_class, scientific_name) %>% 
  nest(occ = ends_with("tude")) 
```


## Compute alpha hull from occ
```{r}
get_ahull_sf <- function(df, alpha = 2){
  ahull <- df %>% 
    ahull(alpha = alpha)  |> 
    hull2spatial::ahull2poly() |> 
  st_as_sf() |>
  st_set_crs(value = 4326)
  
  return(ahull)
}

nest_inv %>% 
  mutate(ahull = map(occ, 
                     get_ahull_sf),
         occ_sf = map(occ, 
                      ~ st_as_sf(.x, coords = c("longitude", "latitude"),
                                            crs = 4326)))  # ERROR ! no slot of name "polygons" for this object of class "SpatialPoints"
```

### Break down the get_ahull_sf function into each step to see where it is failing

```{r}
# Compute ahull is ok! 
nest_inv <- nest_inv %>% 
  mutate(ahull = map(occ,
                     ~ahull(.x, alpha = 2)))

# The error occurred in group 259: family = "Camaenidae", plot_class = "Gastropoda", scientific_name = "Amplirhagada decora".
# Caused by error in `alphahull::ahull()`:
# ! object 'case' not found

# Investigate 259
nest_inv[259,]$occ %>% data.frame() 

# Plot 
nest_inv[259,]$occ %>% data.frame() %>% st_as_sf(coords = c("longitude", "latitude"), crs = 4326) -> occ_sf

ggplot() + 
  geom_sf(data = aus, colour = "black", fill = "white")  +
  geom_sf(data = y, colour = "red", size = 1) + 
  coord_sf(xlim=c(124, 126), 
           ylim=c(-14,-16)) 
  
nest_inv[259,]$occ %>% 
  data.frame() %>% 
  ahull(alpha = 0.01) %>% 
  plot()
  
# Convert to sp object
nest_inv <- nest_inv %>% 
  mutate(ahull_sp = map(ahull,
                        hull2spatial::ahull2poly))

# Convert to sf object
nest_inv %>% 
  mutate(ahull_sf = map(ahull_sp,
                         ~st_as_sf(.x, coords=c("longitude","latitude"), crs = 4326)))

# Caused by error in `slot()`:
# ! no slot of name "polygons" for this object of class "SpatialPoints"

nest_inv$occ[[1]]
nest_inv$ahull[[1]] %>% plot()
nest_inv$ahull_sp[[1]] -> x

plot(x) 
str(x)

x@proj4string <- sp::CRS("EPSG:4326") 
slotNames(x)
x@polygons
                  
?st_sf
?st_sfc

ggplot(data = x, aes(x = longitude, y =  latitude, group = ID)) + 
  #geom_sf(data = aus, colour = "black", fill = "white")  +
  geom_polygon(colour = "red", size = 1) + 
  coord

```

### Trying to spatially thin data

```{r}
test_Dat <- nest_inv[259,]$occ %>% data.frame()

test_Dat <- clean_inv %>% 
  filter(scientific_name == "Amplirhagada decora") %>% 
  select(scientific_name, latitude, longitude) %>%  data.frame()


x <- thin(test_Dat, 
     lat.col = "latitude", 
     long.col = "longitude", 
     spec.col = "scientific_name",
     thin.par = 10, reps = 10,
     locs.thinned.list.return = TRUE,
      write.files = FALSE
      )

plotThin(x)

x
```

```{r}
top5jspiders_ahull %>% 
  select(1:3, ahull) %>% 
  unnest() %>% 
  st_as_sf() -> top5jspiders_subset_ahull

top5jspiders_ahull %>% 
    select(1:3, occ_sf) %>% 
  unnest() %>% 
  st_as_sf() -> top5jspiders_subset_points
```

### PLot spider ahulls

```{r}
library(wesanderson)

wes_palette("Moonrise3")

ggplot() + 
  geom_sf(data = aus, colour = "black", fill = "white")  +
  geom_sf(data = top5jspiders_subset_ahull, aes(fill = scientific_name), alpha = 0.7) +
  geom_sf(data = top5jspiders_subset_points, aes(colour = scientific_name), size = 0.5) +
  scale_fill_manual(values = wes_palette("Moonrise3")) +
  scale_color_manual(values = wes_palette("Moonrise3")) +
  labs(x = "Longtitude (DD)", y = "Latitude (DD)") +
  coord_sf(xlim=c(110, 155), 
           ylim=c(-8,-45)) +
  theme_bw()
```

### Try convex hulls

```{r}
join_chull <-function(chull_vec){
  polygn <- c(chull_vec, chull_vec[1])
  polygn
}

chull_overlap <- function(occ, chull_polygn){
  occ[chull_polygn,]
}

sf_transform <- function(occ){
  st_as_sf() %>% 
  st_set_crs(value = 4326)
}

spatial_inv <- nest_inv %>% 
  filter(scientific_name %in% top5jspiders) %>% 
  mutate(chull = map(occ, 
                     chull),
         chull_polygon = map(chull, 
                             join_chull),
         chull_occ = map2(.x = occ,
                          .y = chull_polygon,
                          chull_overlap)) 


```
### Convert to sf

```{r}
polygon_sf <- function(polygon){
polygon %>%
  st_as_sf(coords=c("longitude","latitude"), crs = 4326) %>%
  summarise(geometry = st_combine(geometry)) %>%
  st_cast("POLYGON") -> x
  
  x
}

spatial_inv <- spatial_inv %>% 
  ungroup %>% 
  mutate(polygon_sf = map(chull_occ, 
                          polygon_sf)) 

spatial_inv$polygon_sf %>%
  bind_rows() %>% 
  mutate(scientific_name = spatial_inv$scientific_name, 
         plot_class = spatial_inv$scientific_name)


spatial_inv %>% 
  select(1:3, polygon_sf) %>% 
  unnest(cols = c(polygon_sf)) %>% 
  sf::st_as_sf(crs = 4326)  -> plot_convex_hull 
```

### Lets try plot it

```{r}
aus <- st_transform(ozmaps::ozmap_country, 4326)

spatial_inv %>%
  mutate(occ_sf = map(occ,
                      ~ st_as_sf(.x, coords = c("longitude", "latitude"),
                                            crs = 4326))) %>% 
  select(scientific_name, occ_sf) %>% 
  unnest() %>% 
  st_as_sf() -> spider_chull_points

ggplot() + 
  geom_sf(data = aus, colour = "black", fill = "white")  +
  geom_sf(data = plot_convex_hull, aes(group = scientific_name, fill = scientific_name), alpha = 0.5) +
  geom_sf(data = spider_chull_points, aes(colour = scientific_name), size = 1) + 
  scale_fill_manual(values = wes_palette("Moonrise3")) +
  scale_color_manual(values = wes_palette("Moonrise3")) +
  labs(x = "Longtitude (DD)", y = "Latitude (DD)") +
  coord_sf(xlim=c(110, 155), 
           ylim=c(-8,-45)) +
  theme_bw() + 
  theme(legend.position = "none")
  
```


### Join in other info

```{r}
plot_convex_hull <- plot_convex_hull %>% 
  left_join(inverts %>% select(scientific_name, family, year), by = "scientific_name")
```

```{r}
# Spiders only
plot_convex_hull %>% 
  filter(plot_class == "Arachnida") %>% 
  group_by(family) %>% 
  summarise(n = length(unique(scientific_name)))  %>% 
  arrange(-n)

# What about Salticidae?! 
plot_convex_hull %>% 
  filter(plot_class == "Arachnida", 
         family == "Salticidae") -> jumping_spiders

ggplot() + 
  geom_sf(data = aus, colour = "black", fill = "white")  +
  geom_sf(data = jumping_spiders, aes(fill = scientific_name), alpha = 0.1) +
  labs(x = "Longtitude (DD)", y = "Latitude (DD)") +
  coord_sf(xlim=c(110, 155), 
           ylim=c(-8,-45)) +
  theme_bw() + 
  theme(legend.position = "none")

# What about the top 5 most species rich taxa
jumping_spiders %>% 
  filter(scientific_name %in% top3jspiders) -> top3jspiders_subset
  
ggplot() + 
  geom_sf(data = aus, colour = "black", fill = "white")  +
  geom_sf(data = top3jspiders_subset, aes(fill = scientific_name), alpha = 0.1) +
  labs(x = "Longtitude (DD)", y = "Latitude (DD)") +
  coord_sf(xlim=c(110, 155), 
           ylim=c(-8,-45)) +
  theme_bw() + 
  theme(legend.position = "none")
```

### Try ConR

```{r}
test_Dat <- test_Dat %>% 
  select(latitude, longitude, scientific_name)
aus_sp <- aus %>% sf::as_Spatial()

IUCN.eval(test_Dat, 
          country_map = aus_sp,
          method.range = "alpha.hull",
          alpha = 2,
          Cell_size_AOO = 2,
          Cell_size_locations = 2,
          exclude.area = TRUE,
          export_shp = TRUE, ## to get SpatialPolygonsDataFrame in output
          write_shp = FALSE, ## to write shapefile files to folder
          SubPop = FALSE,
          DrawMap = FALSE)

# Same error:
#{ : task 1 failed - "task 1 failed - "object 'case' not found""
```

